An OpenCL-based tridiagonal solver for evaluation of compact finite differences on parallel architectures

### 2015-12-01 presentation/background-gpus.tex
updates to presentation

\begin{frame}[t]
\frametitle{Introduction}
\begin{columns}[T]
\begin{column}{0.5\textwidth}
\begin{itemize}
\item Highly multi-threaded processor
    \begin{itemize}
        \item NVIDIA Tesla K20 accelerator: 2496 cores
    \end{itemize}
\item Spectacular performance for
    \emph{compute-intensive, data-parallel} operations
\item Not suited for general-purpose computation
\item Requires careful redesign of algorithms,
    substantial code changes, and tuning efforts
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
\begin{figure}
\begin{center}
\includegraphics[width=150px]{img/device-comparison.png}
\centering
\caption{Design philosophy of GPU (CUDA Programming Guide)}
\end{center}
\end{figure}
\end{column}
\end{columns}
\end{frame}

\begin{frame}
\frametitle{CUDA Programming model}
\begin{itemize}
\item Several options for programming GPUs
    \begin{itemize}[<+->]
        \item underlying models: CUDA and OpenCL
    \end{itemize}
\item Important to understand for good performance
\item Primary features:
    \begin{itemize}
        \item \emph{Kernels} and thread organization
        \item GPU Architecture
        \item Memory hierarchy
    \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Kernels and thread organization}
\begin{columns}
\begin{column}{0.5\textwidth}
\textbf{Kernels}
\begin{itemize}
\item Special pieces of code that execute on the GPU
\item In Fortran: subroutines; in C: functions
\item Executed concurrently by several GPU threads
\end{itemize}
\textbf{Threads}
\begin{itemize}
\item Threads organized into a grid of blocks
\item Kernels launched with specified grid size
    (number of blocks) and block size (threads per block)
\item Grid and blocks can be 2-D or 3-D for convenience
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
\includegraphics[width=120px]{img/program-structure.png}
\vspace{1cm}
\includegraphics[width=120px]{img/grid-of-thread-blocks.png}
\end{column}
\end{columns}
\end{frame}

\begin{frame}
\frametitle{GPU Architecture}
\begin{columns}
\begin{column}{0.5\textwidth}
\begin{itemize}
\item GPU viewed as a collection of \emph{streaming microprocessors} (SMs)
\item When kernel is launched, each block gets assigned to an SM
\item Threads within a block execute concurrently
\item SM may execute several blocks concurrently
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
\includegraphics[width=150px]{img/gpu-scaling.png}
\end{column}
\end{columns}
\end{frame}

\begin{frame}
\frametitle{Memory hierarchy}
\begin{columns}
\begin{column}{0.5\textwidth}
\begin{itemize}
\item Global memory{
\begin{itemize}
    \item large but slow (~5 GB for Tesla K20)
    \item CPU can access (via PCI-e bus)
    \item all threads can access
    \item persists between kernel launches
\end{itemize}}
\item Shared memory{
\begin{itemize}
    \item small but fast (~48 KiB per SM and block)
    \item local to threads within a block
    \item threads in a block read and write into shared memory
    \item explicitly managed cache
\end{itemize}}
\item Registers{
\begin{itemize}
    \item limited (65536 per SM and block)
    \item local to individual threads
    \item fastest
\end{itemize}}
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
    \visible{\includegraphics[width=150px]{img/memory-hierarchy.png}}
\end{column}
\end{columns}
\end{frame}

\begin{frame}
\frametitle{Performance considerations}
\begin{itemize}
\item CPU-GPU transfers
\item Coalesced global memory access
    \begin{itemize}
        \item successive threads must access nearby locations
    \end{itemize}
\item Shared memory and bank conflicts
    \begin{itemize}
        \item \emph{strided} memory access is serialized
    \end{itemize}
\item Limiting use of shared memory and registers
\end{itemize}
\end{frame}

>>>>

\begin{frame}[t]
\frametitle{Graphics processing units}
\begin{columns}[T]
\begin{column}{0.5\textwidth}
\begin{itemize}
\item Highly parallel and multi-threaded accelerator
    \begin{itemize}
        \item NVIDIA Tesla K20 accelerator: 2496 cores
        \item Focus on \emph{throughput}, not execution speed
    \end{itemize}
\item Spectacular performance for workloads that
    are compute-intensive, data-parallel
    (\emph{fine grained parallelism})
\item Effective use requires careful redesign
    of algorithms, code modification,
    and substantial tuning efforts
\item Vendors: NVIDIA, AMD
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
\begin{figure}
\begin{center}
\includegraphics[width=150px]{img/device-comparison.png}
\centering
\end{center}
\end{figure}
\end{column}
\end{columns}
\end{frame}

\begin{frame}
\frametitle{Performance considerations}
\begin{itemize}
\item CPU-GPU transfers expensive -
    hide communication cost by keeping GPU busy
\item Expose as much parallelism as possible
    to utilize GPU cores effectively
\item Avoid thread divergence and excessive
    synchronization
\item Memory access:
    \begin{itemize}
        \item Threads must access memory in \emph{coalesced}
            fashion
        \item Threads must do enough work to hide
            memory access latency
    \end{itemize}
\item Strided or uncoalesced memory access patterns:
    use explicitly managed cache called (\emph{shared} memory)
\end{itemize}
\end{frame}

### 2015-12-01 presentation/compact-finite-difference-application.tex
updates to presentation

\begin{frame}
\frametitle{Application to compact finite difference schemes}
\begin{itemize}
\item Compact finite difference evaluation involves two steps:
    \begin{enumerate}
        \item Evaluation of the right hand sides
        \begin{itemize}
            \item easily performed on the GPU
        \end{itemize}
        \item Solving the tridiagonal systems for all the grid lines
    \end{enumerate}
\item Typical DNS problems too large to fit on a single GPU
\item Need a strategy for multiple GPUs
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Parallelization strategy: single GPU}
\begin{columns}
\begin{column}{0.5\textwidth}
\begin{itemize}
    \item Map grid lines to blocks
    \item Solve multiple blocks simaltaneously
    \item For other coordinate directions,
        transpose the data (physically or logically)
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
\includegraphics[width=150px]{img/compact-single-gpu.eps}
\end{column}
\end{columns}
\end{frame}

@@

\begin{frame}
\frametitle{Parallelization strategy: multiple GPUs on distinct nodes}
\begin{columns}
\begin{column}{0.5\textwidth}
\begin{itemize}
\item Single subdomain used along derivative direction
\item No coordination between GPUs required
\item \textbf{Physical} global transpose required
\item Subdomains can become impractically slender
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
\includegraphics[width=100px]{img/compact-distributed-restricted.eps}
\end{column}
\end{columns}
\end{frame}

\begin{frame}
\frametitle{Parallelization strategy: multiple GPUs on distinct nodes}
\centering
\includegraphics[width=200px]{img/compact-restricted-transpose.eps}
\end{frame}

\begin{frame}
\frametitle{Parallelization strategy: multiple GPUs on distinct nodes}
\begin{columns}
\begin{column}{0.5\textwidth}
\begin{itemize}
\item Need a distributed tridiagonal solver
\item Can ease communication costs
\item Remove restrictions on domain sizes
\item Greatly increases overall complexity
\item Especially difficult for GPUs
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
\centering
\includegraphics[width=100px]{img/compact-distributed-all.eps}
\end{column}
\end{columns}
\end{frame}

@@

\begin{frame}
\begin{itemize}
\item 3 local tridiagonal systems ($m$ equations)
\item 1 ``global'' tridiagonal system ($2P-2$ equations)
\item Must be done for each grid line
\item Sufficient to consider a single line of processes
\end{itemize}
\begin{center}
\includegraphics[height=100px]{img/single-line-of-processes.eps}
\end{center}
\end{frame}

\begin{frame}
\frametitle{Right hand side evaluation}
\begin{columns}
\begin{column}{0.5\textwidth}
\begin{itemize}
\item Stencil pattern: easily performed on GPU
\item Each process communicates with its neighbours
    (halo-swapping)
\item NVIDIA GPUDirect technology for efficient
    GPU-GPU communication
\item Developed a tool to manage structured grid data
    and transparently perform halo swaps
    for multiple GPUs
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
\includegraphics[width=150px]{img/rhs-halos.eps}
\end{column}
\end{columns}
\end{frame}

\begin{frame}
\frametitle{Solving local systems}
\begin{itemize}
\item Local tridiagonal systems solved using
    algorithm described in previous sections
\item For 2-D problems,
    $N_{rhs} = m$
    for 3-D problems
    $N_{rhs} = m^2$
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Constructing reduced system}
\begin{columns}
\begin{column}{0.5\textwidth}
\begin{itemize}
\item Involves global communication
\item Produces tridiagonal system with
    several right-hand sides
\item Systems are \emph{interleaved}
\item Solved by thread-parallel Thomas
    (pThomas) algorithm
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
\includegraphics[width=100px]{img/constructing-reduced-system.eps}
\end{column}
\end{columns}
\end{frame}

\begin{frame}
\frametitle{Summing the solutions}
\begin{itemize}
\item Pointwise-parallel: perfect fit for GPU
\item No communication involved
\end{itemize}
\end{frame}

>>>>

\begin{frame}
\frametitle{Application to compact finite difference schemes}
\begin{itemize}
\item Compact finite difference evaluation involves two steps:
\begin{enumerate}
\item Evaluation of the right hand sides
\begin{itemize}
    \item easily performed on the GPU
\end{itemize}
\item Solving the tridiagonal systems for all the grid lines
\end{enumerate}
\item Current efforts:
\begin{itemize}
    \item Tutkun et al. (2012)
    \item Esfahanian et al. (2014)
\end{itemize}
\item Typical DNS problems too large to fit on a single GPU
\item Need a strategy for multiple GPUs
\end{itemize}
\end{frame}

@@

\begin{frame}
\frametitle{Parallelization strategy: multiple GPUs on distinct nodes}
\begin{columns}
\begin{column}{0.5\textwidth}
\begin{itemize}
\item Single subdomain used along derivative direction
\item No coordination between GPUs required
\item \textbf{Physical} global transpose required
\item Subdomains can become impractically slender
\item Scalable approach needs a distributed tridiagonal solver
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
    \only<1>{\includegraphics[width=100px]{img/compact-distributed-restricted.eps}}
    \only<2>{\includegraphics[width=150px]{img/compact-restricted-transpose.eps}}
    \only<3>{\includegraphics[width=120px]{img/compact-distributed-all.eps}}
\end{column}
\end{columns}
\end{frame}

@@

\begin{frame}
\frametitle{Compact finite difference on multi-GPU system}
\footnotesize
\begin{columns}
\begin{column}{0.5\textwidth}
\begin{enumerate}
\item Computing RHS for all local grid points - \textbf{requires communication}
\begin{itemize}
    \footnotesize
    \item halo swaps with neighbouring GPUs
    \item pointwise stencil
\end{itemize}
\item Solving local tridiagonal systems - no communication
\item Assembling and solving reduced tridiagonal systems - \textbf{requires communication}
\begin{itemize}
    \footnotesize
    \item global communication along line of GPUs
    \item produces interleaved tridiagonal sysems - pThomas algorithm
\end{itemize}
\item Summing solutions from steps 2 and 3 - no communication
\begin{itemize}
    \footnotesize
    \item pointwise sum
\end{itemize}
\item For other coordinate directions: \emph{local} permutation of data
\end{enumerate}
\end{column}
\begin{column}{0.5\textwidth}
\only<1>{\includegraphics[width=150px]{img/rhs-halos.eps}}
\only<2>{\includegraphics[width=150px]{img/constructing-reduced-system.eps}}
\end{column}
\end{columns}
\end{frame}

### 2015-12-01 presentation/results.tex
updates to presentation

\begin{frame}
\frametitle{Overview of results}
\begin{itemize}
\item Results for tridiagonal solver
\begin{itemize}
    \item Global memory v/s shared memory implementations
    \item Compared against library solvers
        (multicore CPU and GPU)
\end{itemize}
\item Results for compact finite difference application
\begin{itemize}
    \item Profiling results
    \item Scaling results
    \item Comparison with CPU-only approach
\end{itemize}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Global memory v/s shared memory implementations}
\begin{columns}
\begin{column}{0.5\textwidth}
\begin{itemize}
\item For 2-D problems, shared memory clearly better
\item For 3-D problems, shared memory speedup
    is smaller
\item Global memory makes better use of
    GPU parallelism in 3-D case
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
\centering
\includegraphics[width=140px]{img/global-vs-shared-2d.eps}

@@

\begin{frame}
\frametitle{Comparison with library solvers - timings}
\begin{table}
\resizebox{\textwidth}{!}{%
\input{table/bench.tex}
}
\end{table}
\end{frame}

\begin{frame}
\frametitle{Comparison with library solvers - speedup}
\centering
\includegraphics[width=180px]{img/bench-2d.eps}
\includegraphics[width=180px]{img/bench-3d.eps}
\begin{table}
\end{table}
\end{frame}

\begin{frame}
\frametitle{Note on speedups}
\begin{itemize}
\item Comparing against 8 CPU cores, speedups in the range 1-2x
\item Solving the tridiagonal systems
    is the \emph{least amenable to efficient parallel solution}
\item Objective is to keep the problem on the GPU
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Compact finite difference evaluation - profiling}
\begin{columns}
\begin{column}{0.5\textwidth}
\begin{itemize}
    \item Problem sizes: $1024^3$ and $2048^3$ on 64 GPUs
    \item For larger problems, more time spent
        on tridiagonal systems
    \item Significant portion of time spent on
        data permutation
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
\centering
\includegraphics[width=170px]{img/profiling-1024-64.eps}

@@

\begin{frame}
\frametitle{Comparison with CPU-only approach}
\begin{columns}
\begin{column}{0.5\textwidth}
\begin{itemize}
\item Reference implementation: approach in CFDNS code
    (Los Alamos National Lab)
\item Maintain a ratio of 8 CPU cores : 1 GPU in comparison
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
\centering
\begin{table}
\resizebox{0.75\textwidth}{!}{%
\input{table/compact-refimpl-timings.tex}
}
\end{table}

@@

>>>>

\begin{frame}
\frametitle{Overview of results}
\begin{itemize}
\item Results for tridiagonal solver (NEATO)
\begin{itemize}
    \item Global memory v/s shared memory implementations
    \item Compared against library solvers
        (multicore CPU and GPU)
\end{itemize}
\item Results for compact finite difference application
\begin{itemize}
    \item Profiling results
    \item Scaling results
    \item Comparison with CPU-only approach
\end{itemize}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Global memory v/s shared memory implementations}
\footnotesize
\begin{columns}
\begin{column}{0.5\textwidth}

\begin{block}{Benchmark details}
\begin{itemize}
\item Solving multiple tridiagonal systems on single GPU
\item Upto $2048$ equations in $2048$ variables (2-D)
    and $512^2$ equations in $512$ variables (3-D)
\item GPU: NVIDIA Tesla K20
\end{itemize}
\end{block}

\begin{itemize}
\item For 2-D problems, shared memory clearly better
\item For 3-D problems, shared memory speedup
    is smaller
\item Thread inactivity in shared memory leads
    to reduced performance
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
\centering
\includegraphics[width=140px]{img/global-vs-shared-2d.eps}

@@

\begin{frame}{Comparison with library solvers}
\footnotesize
\begin{block}{Benchmark details}
\begin{itemize}
\item Solving multiple tridiagonal systems on single GPU and multicore CPU
\item Upto $2048$ equations in $2048$ variables (2-D)
    and $512^2$ equations in $512$ variables (3-D)
\item GPU: NVIDIA Tesla K20
\item CPU: Intel Xeon E5 v2
\item Do \emph{not} include cost of CPU-GPU transfer
\item Library solver details:
\begin{itemize}
    \footnotesize
    \item Intel MKL \texttt{dgtsv} (Gaussian elimination with partial pivoting)
    \item CUSPARSE \texttt{dgtsv} (Hybrid CR+PCR)
\end{itemize}
\item \texttt{-02} level compiler optimizations
\item Double precision
\end{itemize}
\end{block}
\end{frame}

\begin{frame}
\frametitle{Comparison with library solvers - speedup}
Speedups over library implementations:
\begin{center}
\includegraphics[width=180px]{img/bench-2d.eps}
\includegraphics[width=180px]{img/bench-3d.eps}
\end{center}
\end{frame}

\begin{frame}
\frametitle{Comparison with library solvers - timings}
Timings for various solvers:
\begin{table}
\resizebox{\textwidth}{!}{%
\input{table/bench.tex}
}
\end{table}
\end{frame}

\begin{frame}
\frametitle{Note on speedups}
\begin{itemize}
\item Faster than CUSPARSE in almost all cases
\item Comparing against 8 CPU cores,
    speedups over MKL in the range 1-2x
\item Solving the tridiagonal systems
    \emph{least amenable to efficient parallel solution}
\item Objective is to keep the problem on the GPU
\item Multicore CPU performance flattens out
\end{itemize}
\end{frame}

\begin{frame}{Compact finite difference evaluation}
\footnotesize
\begin{block}{Benchmark details}
\begin{itemize}
\item Compact finite difference evaluation in 3-D for a function $f(x, y, z)$
\item Problem size up to $2048^3$ on up to 64 GPUs and 512 CPU cores
\item Tests performed on Palmetto cluster (Infiniband interconnect)
\end{itemize}
\end{block}
\end{frame}

\begin{frame}
\frametitle{Compact finite difference evaluation - profiling}
\begin{columns}
\begin{column}{0.5\textwidth}
\begin{itemize}
\item Problem sizes: $1024^3$ and $2048^3$ on 64 GPUs
\item Most time spent on solving tridiagonal systems
    \begin{itemize}
        \item Justifies optimization effort
    \end{itemize}
\item Significant portion of time spent on
    data permutation
    \begin{itemize}
        \item na\"{\i}ve implementation
    \end{itemize}
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
\centering
\includegraphics[width=170px]{img/profiling-1024-64.eps}

@@

\begin{frame}
\frametitle{Comparison with CPU-only approach}
\begin{columns}
\begin{column}{0.5\textwidth}
\begin{itemize}
\item Reference implementation: CFDNS
    (Los Alamos National Lab)
\item Uses LU algorithm
\item Parallelized among individual cores
\item Maintain a ratio of 8 CPU cores : 1 GPU for comparison
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
\centering
\begin{table}
\resizebox{0.75\textwidth}{!}{%
\input{table/compact-refimpl-timings.tex}
}
\end{table}

@@

\begin{frame}
\frametitle{Conclusions and future work}
\begin{itemize}
    \item Tridiagonal solver performance
        is key in compact finite difference evaluation
    \item Simple matrix structure can be exploited
        in the cyclic reduction algorithm
    \item Substantial speedups can be achieved
        over CPU
    \item Applicability to other algorithms: PCR, CR+PCR
    \item Integration into current DNS code
    \item Strategies for MIC architecures
\end{itemize}
\end{frame}

\begin{frame}
    Thank you    
\end{frame}
